---
title: "Bootstrap"
author: "Armando Teixeira-Pinto"
date:  "`r Sys.Date()`"
output: 
  html_document

---


# Bootstrap

  Bootstrap was proposed by 
  [Efron in 1979](https://projecteuclid.org/download/pdf_1/euclid.aos/1176344552) 
  and it mimics the sampling process by sampling with replacement 
  the original sample. This process replicates the sample variation and allows
  the calculation of standard errors. It can also be used to refine more
  complex machine learning algorithms, as we will see later.
  
  From the original sample of size $n$, we create many (e.g. 10,000) sample of
  size $n$ by sampling the sample with replacement. Notice that the replacement
  allows the same value to be included multiple times in the same sample. This 
  is why you can create many different samples.
  For each of the bootstrapped sample we compute the statistics of interest. The
  standard deviance of all these computed statistics, is the standard error.
  
  
 <iframe width="560" height="315" src="https://www.youtube.com/embed/U8r-UrwTOy4" 
 frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; 
 picture-in-picture" allowfullscreen></iframe>

## Readings

Read the following chapter of *An introduction to statistical learning*: 

* 5.2 The Bootstrap


## Practical session


### Task - Confidence intervals with bootstrap {-}

We will be using the 
[bmd.csv](https://www.dropbox.com/s/7wjsfdaf0wt2kg2/bmd.csv?dl=1) 
dataset to plot the histogram, compute the median and 95% confidence interval 
for the "waiting_time"

Let's first read the data and create the variable BMI that will be used later.
```{r message = FALSE }
library(caret) #library for Machine Learning
library(pROC)  #library for the ROC curve
set.seed(1974)

#the option stringsAsFactors = TRUE in the command below converts 
#string variables as sex into factor variables
bmd.data <- read.csv("https://www.dropbox.com/s/c6mhgatkotuze8o/bmd.csv?dl=1", 
                     stringsAsFactors = TRUE)
#computes the BMI
bmd.data$bmi <- bmd.data$weight_kg / (bmd.data$height_cm/100)^2
```

Let's plot the histogram for *waiting time* and compute the median
```{r}
hist(bmd.data$waiting_time)
median(bmd.data$waiting_time)

```


And now use the function **boot()**, from the library **boot**, to bootstrap the median. We will also compute it manually

```{r}
#using the boot function
median(bmd.data$waiting_time)
samplemedian <- function(x, d) { #need to define the function to use bootstrap
  return(median(x[d]))           #d is the index for the bootstrap
}
bootresults <- boot(bmd.data$waiting_time, statistic=samplemedian, R=10000)
# get 95% confidence interval 
boot.ci(bootresults, type="perc")

#manual bootstrap
median.bs <- NA
for (i in 1:10000){ # change to 10000
  sample.bs    <- sample(bmd.data$waiting_time, 169, replace = TRUE)
  median.bs[i] <- median(sample.bs)
}
median(median.bs)  #median (could use mean) of all the bootstrapped medians
sd(median.bs)      #Std error of the median
quantile(median.bs, c(0.025, 0.975))  #95% empirical confidence interval
```

**TRY IT YOURSELF:**

1) Compute the mean for *waiting_time* and the usual 95% confidence 
interval using the **CI()** function.

<details><summary>*See the solution code*</summary>

```{r results="hide"}
#library(Rmisc) #CI() function to compute the conf interval for the mean
CI(bmd.data$waiting_time)
```
</details><p><p>    

2) Compute the 95% confidence interval for the mean using the boot function

<details><summary>*See the solution code*</summary>

```{r results="hide"}
CI(bmd.data$waiting_time)

samplemean <- function(x, d) { #need to define the function to use bootstrap
  return(mean(x[d]))           #d is the index for the bootstrap
}

bootresults <- boot(bmd.data$waiting_time, statistic=samplemean, R=10000)
boot.ci(bootresults, type="perc")
```
</details><p><p>




## Exercises

Solve the following exercises:

The diabetes data were provided by Hastie and Tibshirani (1990, p. 6). 
The observations arise from a study of the factors affecting patterns of 
insulin-dependent diabetes mellitus in 43 children (Sockett et al., 1987). 
The aim was to investigate the dependence of serum C-peptide on other factors, 
better to understand the patterns of residual insulin secretion. 

The response, *cpep*, is the log of C-peptide concentration at diagnosis, and 
the selected covariates are *age*, the child’s age at diagnosis, and *base*, 
minus their base deficit. Base deficit is a measure of acidity. 

**1) Calculate the Pearson correlation coefficient between *cpep* and *base* with
the respective 95% confidence interval obtained by the [Fisher's 
z-transformation](https://en.wikipedia.org/wiki/Fisher_transformation) (the 
usual way of getting the confidence interval for the Pearson's correlation)**

*Note: you can use the function CIr in the "psychometric" package*

**2) Write your own function to compute the 95% confidence interval 
for the above correlation, using bootstrap.**

**3) Use the function *boot()* from the *boot* package to compute 
the 95% confidence interval through bootstrap**

**4) Plot the histogram with the correlations obtained in 
the bootstrap samples.**
 
<details><summary>*See the solution code*</summary>

```{r boostrapexercise, results="hide", message=FALSE}
#install.packages("psychometric") # install package with function for
                                 # the correlation confidence interval
#install.packages("boot")         # install package for bootstrap function
library(boot)
library(psychometric) 
set.seed(1001)
myData <- read.csv("https://www.dropbox.com/s/6rc00ealjtyp3qi/diabetes.csv?dl=1")

sample.size <- dim(myData)[1] #nr of observations

#1 - the Person correlation
  cor(myData$base, myData$cpep)
  CIr( r = cor(myData$base, myData$cpep),   #conf interval using
       n = sample.size, level = .95)        # Fisher's z-transformations

#2 - Bootstrap
  n.boot   <- 25000  #choose how many bootstraps
  cor.boot <- NULL # to store the bootstrap correlations
  
  #manually implementing bootstrap
  for (i in 1:n.boot) {
    id.bs <- sample.int(sample.size,   #bootstrap the
                        sample.size,   #original sample
                        replace = TRUE)
    cor.boot[i] <- cor(myData[id.bs, ])[2,3]  #Compute correlation
                                              #between
   }
   quantile(cor.boot, c(0.025, 0.975))  #95% confidence interval
  
#3 - Using the boot() function
  sample.corr <- function(data, d) {  
    return(cor(data$base[d], data$cpep[d]))   #d is the index for the bootstrap
  }
    
  bootcorr <- boot(myData, 
                   statistic=sample.corr, 
                   R=n.boot)
  # get 95% confidence interval 
  boot.ci(bootcorr, type="perc")

#4 - Histogram
  #the correlations for the bootstrap samples
  #are stored in bootcorr$t
  hist(bootcorr$t)
  
  #or 
  #hist(cor.boot)
```
</details>  
