---
title: "Cross-validation"
author: "Armando Teixeira-Pinto"
date:  "`r Sys.Date()`"
output: 
  html_document

---


# Cross-validation

  When assessing the performance of a model in the same data that was used to 
  fit the model, we will be overestimating the model performance. A better 
  strategy is to initially split the data and use one part to fit the model and
  the other one to test it. In machine learning terminology the data used to fit
  the model is called the **training data** and the data used to assess the model 
  performance is called **testing data**.
  
  The cross-validation is a repetition of the process above but each time we 
  use a different split of the data. This will result in several measures of
  performance obtained in each split combination. The final performance 
  statistics is obtained by averaging all results of the different splits.
  
 
  
  
 <iframe width="560" height="315" src="https://www.youtube.com/embed/QrcYxR5fd-w" 
 frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; 
 picture-in-picture" allowfullscreen></iframe>

## Readings

Read the following chapter of *An introduction to statistical learning*: 

* 5.1 Cross-validation


## Practical session


### Task 1 - Cross-validated MSE and R^2 {-}

We will be using the 
[bmd.csv](https://www.dropbox.com/s/7wjsfdaf0wt2kg2/bmd.csv?dl=1) 
dataset to fit a linear model for *bmd* using *age*, *sex* and *bmi*, 
and compute the cross-validated MSE and $R^2$.
We will fit the model with main effects using 10 times a 5-fold cross-validation. 

We will use the tools from the **caret** package. 
This is a powerful package that wraps several methods for regression and 
classification: [manual](http://topepo.github.io/caret/index.html)

<iframe width="560" height="315" src="https://www.youtube.com/embed/1FuZCTjNPDU" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; 
picture-in-picture" allowfullscreen></iframe>



```{r message = FALSE }
library(caret) #library for Machine Learning
library(boot)  #library for bootstrap
library(pROC)  #library for the ROC curve
library(Rmisc) #CI() function to compute the conf interval for the mean
set.seed(1974)
#the option stringsAsFactors = TRUE in the command below converts 
#string variables as sex into factor variables
bmd.data <- read.csv("https://www.dropbox.com/s/c6mhgatkotuze8o/bmd.csv?dl=1", 
                     stringsAsFactors = TRUE)
#computes the BMI
bmd.data$bmi <- bmd.data$weight_kg / (bmd.data$height_cm/100)^2
```


```{r}
trC.lm <- trainControl(                  #defines the CV procedure
                 method = "repeatedcv",  #multiple CV
                 number = 5,            #5-fold CV
                 repeats = 10)          #repeats the cross validation 10 times

#fits the linear model with CV defined above
model.lm <- train(bmd ~ age + sex + bmi, 
                data = bmd.data, 
                method = "lm", 
                trControl = trC.lm)

model.lm
summary(model.lm$finalModel)
```

**TRY IT YOURSELF:**

1) Refit the same model and evaluate the MSE and $R^2$ using leave one out (**method = "LOOCV"**)

<details><summary>*See the solution code*</summary>

```{r results="hide"}
trC.lm.loocv <- trainControl(  
                 method = "LOOCV") # leave on out CV
                   
model.lm2 <- train(bmd ~ age + sex + bmi, 
                data = bmd.data, method = "lm",
                trControl = trC.lm.loocv)

model.lm2
summary(model.lm2$finalModel)
```

</details><p><p>  

2)  Fit a k nearest neighbour regression for BMD using AGE, SEX and BMI, and choose the k (number of neighbours) by 10-fold cross-validation repeated 10 times. Also, obtain the MSE and  $R^2$

<details><summary>*See the solution code*</summary>

```{r results="hide"} 
trC.knn <- trainControl(    
                 method = "repeatedcv", # multiple CV
                 number = 10,  #10-fold CV
                 repeats = 10)

model.knn <- train(bmd ~ age + sex + bmi, 
                data = bmd.data, method = "knn",
                trControl = trC.knn,
                tuneLength = 20)

# Model Summary
model.knn
model.knn$results

#results in each cv-fold
model.knn$resample
plot(model.knn)

```
</details><p><p>  

### TASK 2 - ROC cross-validation {-}

We want to fit the following model **logit(fracture) ~ age + sex + bmi +bmd** 
and assess its performance by computing the area under the ROC curve 
using cross-validation

```{r}
#caret does not like the category "no fracture"
#because of the space
#We are creating a new label for the categories
bmd.data$fract <- ifelse (bmd.data$fracture =="fracture", "F", "NF")
bmd.data$fract <- as.factor(bmd.data$fract)

trC <- trainControl(
            method = "cv",     #just 1 CV
            number = 10,       #10-fold CV
            classProbs = TRUE,
            summaryFunction = twoClassSummary,
            savePred =TRUE)  #to be used in the confusion matrix

model.LR <- train(fract ~ age + sex + bmi + bmd , 
                data = bmd.data, method = "glm",
                family="binomial",
                trControl = trC,
                metric = "ROC")


# Model Summary
model.LR
model.LR$results

#results in each cv-fold
model.LR$resample

#Confusion matrix cross-validated
confusionMatrix(model.LR)

#Confusion matrix for the final model
pred.LR  <- predict(model.LR)
confusionMatrix(data=pred.LR, reference=bmd.data$fract)

```

**TRY IT YOURSELF**

1) Use the KNN algorithm to predict **fracture** based on the same variables of the logistic model above, by choosing the *k* using cross-validation and compute the area under the ROC.


```{r echo=FALSE}
trC.knn <- trainControl(
            method = "cv",     #just 1 CV
            number = 10,       #10-fold CV
            classProbs = TRUE,
            summaryFunction = twoClassSummary,
            savePred =TRUE)  #to be used in the confusion matrix

model.knn <- train(fract ~ age + sex + bmi + bmd , 
                data = bmd.data, method = "knn",
                trControl = trC.knn,
                tuneLength = 20,
                metric = "ROC")


# Model Summary
model.knn
plot(model.knn)
model.knn$results

#results in each cv-fold
model.knn$resample

#Confusion matrix cross-validated
confusionMatrix(model.knn)

#Confusion matrix for the final model
pred.knn  <- predict(model.knn)
confusionMatrix(data=pred.knn, reference=bmd.data$fract)

```



### TASK 3 - Classification

Read the dataset **SBI.csv** in and create a prediction model for the outcome "sbi" using "age", "pct", "crp", "wcc" and "fever_hours" as predictors 
Let's first read the data in
```{r}
sbi.data     <- 
  read.csv("https://www.dropbox.com/s/wg32uj43fsy9yvd/SBI.csv?dl=1", 
           stringsAsFactors = TRUE)
summary(sbi.data)
```

We will try different approaches starting with linear discriminant analysis

```{r}
trCtrl.lda <- trainControl(method = "repeatedcv",
                           number = 10,  #10-fold CV
                           repeats = 10,
                           classProbs = TRUE,
                           savePredictions = TRUE)
model.lda <- train(sbi ~ crp+pct+age+wcc+fever_hours, 
                   data=sbi.data, 
                   method="lda",
                   trControl = trCtrl.lda,
                   metric="Accuracy" )

model.lda$results
confusionMatrix(predict(model.lda), sbi.data$sbi)
```

Now, let's try logistic regression
```{r}
trCtrl.lr <- trainControl(method = "repeatedcv",
                           number = 10,  #10-fold CV
                           repeats = 10,
                           classProbs = TRUE,
                           savePredictions = TRUE)
model.lr <- train(sbi ~ crp+pct+age+wcc+fever_hours, 
                   data=sbi.data, 
                   method="multinom",
                   trControl = trCtrl.lr,
                   tuneLength=1)

model.lr$results
confusionMatrix(predict(model.lr), sbi.data$sbi)
```

And finally, knn.

```{r}
trCtrl.knn <- trainControl(method = "repeatedcv",
                           number = 10,  #10-fold CV
                           repeats = 10,
                           classProbs = TRUE,
                           savePredictions = TRUE)
model.knn <- train(sbi ~ crp+pct+age+wcc+fever_hours, 
                   data=sbi.data, 
                   method="knn",
                   trControl = trCtrl.knn,
                   tuneLength=20)

plot(model.knn)
model.knn
confusionMatrix(predict(model.knn), sbi.data$sbi)
```






## Exercises

Solve the following exercises:

The diabetes data were provided by Hastie and Tibshirani (1990, p. 6). 
The observations arise from a study of the factors affecting patterns of 
insulin-dependent diabetes mellitus in 43 children (Sockett et al., 1987). 
The aim was to investigate the dependence of serum C-peptide on other factors, 
better to understand the patterns of residual insulin secretion. 

The response, *cpep*, is the log of C-peptide concentration at diagnosis, and 
the selected covariates are *age*, the child’s age at diagnosis, and *base*, 
minus their base deficit. Base deficit is a measure of acidity. 

**1) Calculate the Pearson correlation coefficient between *cpep* and *base* with
the respective 95% confidence interval obtained by the [Fisher's 
z-transformation](https://en.wikipedia.org/wiki/Fisher_transformation) (the 
usual way of getting the confidence interval for the Pearson's correlation)**

*Note: you can use the function CIr in the "psychometric" package*

**2) Write your own function to compute the 95% confidence interval 
for the above correlation, using bootstrap.**

**3) Use the function *boot()* from the *boot* package to compute 
the 95% confidence interval through bootstrap**

**4) Plot the histogram with the correlations obtained in 
the bootstrap samples.**
 
<details><summary>*See the solution code*</summary>

```{r boostrapexercise, results="hide", message=FALSE}
#install.packages("psychometric") # install package with function for
                                 # the correlation confidence interval
#install.packages("boot")         # install package for bootstrap function
library(boot)
library(psychometric) 
set.seed(1001)
myData <- read.csv("https://www.dropbox.com/s/6rc00ealjtyp3qi/diabetes.csv?dl=1")

sample.size <- dim(myData)[1] #nr of observations

#1 - the Person correlation
  cor(myData$base, myData$cpep)
  CIr( r = cor(myData$base, myData$cpep),   #conf interval using
       n = sample.size, level = .95)        # Fisher's z-transformations

#2 - Bootstrap
  n.boot   <- 25000  #choose how many bootstraps
  cor.boot <- NULL # to store the bootstrap correlations
  
  #manually implementing bootstrap
  for (i in 1:n.boot) {
    id.bs <- sample.int(sample.size,   #bootstrap the
                        sample.size,   #original sample
                        replace = TRUE)
    cor.boot[i] <- cor(myData[id.bs, ])[2,3]  #Compute correlation
                                              #between
   }
   quantile(cor.boot, c(0.025, 0.975))  #95% confidence interval
  
#3 - Using the boot() function
  sample.corr <- function(data, d) {  
    return(cor(data$base[d], data$cpep[d]))   #d is the index for the bootstrap
  }
    
  bootcorr <- boot(myData, 
                   statistic=sample.corr, 
                   R=n.boot)
  # get 95% confidence interval 
  boot.ci(bootcorr, type="perc")

#4 - Histogram
  #the correlations for the bootstrap samples
  #are stored in bootcorr$t
  hist(bootcorr$t)
  
  #or 
  #hist(cor.boot)
```
</details>  
